---
title: "Homework 5"
author: "CS 498, Spring 2018, Xiaoming Ji"
date: ''
output:
  html_document:
    toc: yes
  pdf_document: default
params:
  load_precomputed_data: yes
---


#Problem 1
At http://www.statsci.org/data/general/brunhild.html, you will find a dataset that measures the concentration of a sulfate in the blood of a baboon named Brunhilda as a function of time. Build a linear regression of the log of the concentration against the log of time.

##(a) 
Prepare a plot showing (a) the data points and (b) the regression line in log-log coordinates.

```{r}
brunhild = read.delim("brunhild.txt")
model = lm(log(Sulfate) ~ log(Hours), data = brunhild)
plot(log(brunhild$Hours), log(brunhild$Sulfate), xlab = "LOGHOURS", ylab = "LOGSULFATE")
abline(model)
```

##(b) 
Prepare a plot showing (a) the data points and (b) the regression curve in the original coordinates.
```{r}
plot(brunhild)
```

##(c) 
Plot the residual against the fitted values in log-log and in original coordinates.

##(d) 
Use your plots to explain whether your regression is good or bad and why.

#Problem 2
At http://www.statsci.org/data/oz/physical.html, youwillfindadatasetofmea- surements by M. Larner, made in 1996. These measurements include body mass, and various diameters. Build a linear regression of predicting the body mass from these diameters.

##(a) 
Plot the residual against the fitted values for your regression.

##(b) 
Now regress the cube root of mass against these diameters. Plot the residual against the fitted values in both these cube root coordinates and in the original coordinates.

##(c) 
Use your plots to explain which regression is better.

#Problem 3
At https://archive.ics.uci.edu/ml/datasets/Abalone, you will find a dataset of measurements by W. J. Nash, T. L. Sellers, S. R. Talbot, A. J. Cawthorn and W. B. Ford, made in 1992. These are a variety of measurements of blacklip abalone (Haliotis rubra; delicious by repute) of various ages and genders.
##(a) 
Build a linear regression predicting the age from the measurements, ignoring gender. Plot the residual against the fitted values.

##(b) 
Build a linear regression predicting the age from the measurements, including gender. There are three levels for gender; I’m not sure whether this has to do with abalone biology or difficulty in determining gender. You can represent gender numerically by choosing 1 for one level, 0 for another, and -1 for the third. Plot the residual against the fitted values.

##(c) 
Now build a linear regression predicting the log of age from the measurements, ignoring gender. Plot the residual against the fitted values.

##(d) 
Now build a linear regression predicting the log age from the measurements, including gender, represented as above. Plot the residual against the fitted values.

##(e) 
Itturnsoutthatdeterminingtheageofanabaloneispossible,butdifficult (you section the shell, and count rings). Use your plots to explain which regression you would use to replace this procedure, and why.

##(f) 
Can you improve these regressions by using a regularizer? Use glmnet to obtain plots of the cross-validated prediction error.
